Higher-Order unification for free!

### 2024-04-29 paper/main.tex
proofread section 2

\subsubsection{Term equality: \Eo and \Ee}
For both languages, we extend the equational theory
over ground terms to the full language by adding the reflexivity for
unification variables (a variable is equal to itself).

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail,
that \elpiIn{forall2} applies a binary predicate to two lists (of same
length). Finally \elpiIn{=>} is $\lambda$Prolog syntax to augment the
current program with an extra rule, and both procedures use it to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity, we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.

@@

\noindent
We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
reflected by the signature above that relates two substitutions.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every higher-order variable is applied to a list of distinct names.

@@

Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of \emph{steps} of length $n$.
At each step $p$ we unify two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_n$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{n}{\rho}
\deff
\runFD{\foUnifPb}{n}
\end{array}
$$

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable map $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

We simulate each run in \Fo{} with a run in \Ho as follows:
%
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{n}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{n}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{n}}{\mapStore}{\linkStore_{n}}{\rho_{n}}
\end{array}
$$

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{~ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} ~\}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honored by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
finally decompiles the solution.
We claim:

@@

In other words, if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
is that the assumption of the former is purely static, it can be checked upfront. 
When this assumption is not satisfied, one can still simulate a logic program and
have guarantees of fidelity if, at run time, decidability of higher-order
unification is restored.

@@

\paragraph{Backtracking} We omit it from our model of a logic-program execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with (possibly non-empty) common prefixes.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' subterm $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but they cannot be handled directly by the unification of the meta language, which
is only required to handle terms in \llambda. Still, there exists a
substitution $\rho$ such that $\rho s \in \llambda$.

@@

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
We write $\subterm{t}$ the set of sub-terms of $t$, and
%\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
%  largest set $\subterm{t}$ that can be obtained by the following rules.
%$$
%\begin{array}{l}
%t \in \subterm{t}\\
%t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
%t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
%\end{array}
%$$
%\end{definition}
%\noindent
we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.

\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X},
t \not\in (\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{})
$$
\end{definition}

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not. And if two terms can unify in the source language
then these terms also unify in the target language. 
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
\Ue{} does not contradict \Eo. Also note that a compiler that
translate all terms to unification variables would satisfy this
property, but wound fail to verify \cref{prop:fidelity}.

@@

We prove \fullref{prop:fidelity} in
\cref{sec:eta,sec:beta,sec:invariant1}
and \fullref{prop:simulation} only in \cref{sec:llam},
where links are not present yet.
In order to prove \cref{prop:simulation} in the presence of links
one needs to refine \cref{prop:fidelity} by relating the two
substitutions: they are equivalent only by using decompilation to 
take into account terms in \linkStore.
Then, by induction
on the number of steps, \ref{prop:fidelity} implies \Cref{prop:simulation}.
We discuss \fullref{prop:fidelity-recovery} in \cref{sec:beta}.

>>>>

\subsubsection{Term equality: \Eo and \Ee}
For both languages, we extend the equational theory
over ground terms to the full language by adding reflexivity for
unification variables, i.e. a unification variable is equal to itself.

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail;
that \elpiIn{forall2} applies a binary predicate to two lists (of same
length); and that \elpiIn{=>} is $\lambda$Prolog syntax for
implication an is used to augment the
current program with an extra rule. Both procedures use implication to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity, we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak-head normal form of <<\elpiIn{fapp [F|L]}>>.

@@

\noindent
We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
reflected by the signature above that relates two substitutions.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every unification variable is applied to a list of distinct names.

@@

Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is to simulate the execution of
an entire logic-program.

\subsection{The problem: logic-program simulation}
We represent a logic program \emph{run} in \Fo{} as
a sequence of $n$ \emph{steps}. Each step consists in
unifying two terms, $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$, taken from the list of all unification
problems \foUnifPb.
% \footnote{If the same rule is used multiple times in a run we
% consider as many copies as needed of the terms composing the
% rules, with fresh unification variables each time. Also, we do not
% model backtracking directly, but rather the fact that a step in a run may
% fail, and in that case the entire run is said to fail. A regular logic program
% that succeeds is modeled as a many runs, with possibly overlapping starts,
% where at least one run does not end in a failure. }
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_n$, which is the result of the
logic program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{ostep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{mstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{orun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{mrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFx}[3]{\ensuremath{\frun(#1,#2) \mapsto #3}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHx}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{n}{\rho}
\deff
\runFD{\foUnifPb}{n}
\end{array}
$$

\noindent
In order to simulate a logic program we start by compiling
\foUnifPb to \hoUnifPb, i.e.
each \Fo{}-term $s \in \foUnifPb$ is translated to a \Ho{}-term $t \in \hoUnifPb$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable map $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l},\, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

Then we simulate each run in \Fo{} with a run in \Ho as follows:
%
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
  \runH{\foUnifPb}{n}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t,m,l) | s \in \foUnifPb, \C{s}{t}{m}{l} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{n}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{n}}{\mapStore}{\linkStore_{n}}{\rho_{n}}
\end{array}
$$

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{~ \hoUnifPb_{p_l},\, \hoUnifPb_{p_r} ~\}$.\\
The \hstep{} procedure is made of two sub-steps: a call to the meta-language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honored by \Ue.
Procedure \hrun{} compiles all terms in \foUnifPb{}, then executes each step, and
finally decompiles the solution.
We claim:

@@

\noindent
In other words, if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.
Indeed, the main difference between \cref{prop:fidelity} and \cref{prop:fidelity-recovery}
is that the assumption of the former is purely static, it can be checked upfront. 
When this assumption is not satisfied, one can still simulate a logic program and
have guarantees of fidelity if, at run time, decidability of higher-order
unification is restored.

@@

\paragraph{Backtracking} We omit it from our model of a logic-program execution
since it plays a very minor role, orthogonal to higher-order unification.
We point out that each \emph{run} corresponds to a (proof search) branch in the
logic program that either fails at some point, or succeeds. A computation that
succeeds by backtracking, exploring multiple branches, could be
modeled as a set of runs with a (possibly non-empty) common prefix.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' subterm $e$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue e$. As a result, \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally
We use the $\Diamond$ symbol since it stands for ``possibly'' in modal logic
and all problematic terms are characterized by some ``uncertainty''.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta-long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but they cannot be handled directly by the unification of the meta language, which
is only required to handle terms in \llambda. Still, for a $s \in \notllambda$
there may exist a substitution $\rho$ such that $\rho s \in \llambda$.

@@

\newcommand{\subterm}[1]{\ensuremath{\mathcal{P}(#1)}}
We write $\subterm{t}$ for the set of sub-terms of $t$, and
%\begin{definition}[Subterms \subterm{t}] The set of sub terms of $t$ is the
%  largest set $\subterm{t}$ that can be obtained by the following rules.
%$$
%\begin{array}{l}
%t \in \subterm{t}\\
%t = f t_1\ldots t_n \Rightarrow \subterm{t_i} \subseteq \subterm{t} \land f \in \subterm{t}\\
%t = \lambda x.t' \Rightarrow \subterm{t'} \subseteq \subterm{t}\\
%\end{array}
%$$
%\end{definition}
%\noindent
we write $\subterm{X} = \bigcup_{t\in X} \subterm{t}$ when $X$ is a set of terms.

\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
\begin{definition}[Well behaved set]
Given a set of terms $X$, % \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X},
t \not\in (\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{})
$$
\end{definition}

@@

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not. And if two terms can unify in the source language
then their images also unify in the target language. 
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: when applied to terms in \wellb,
\Ue{} does not contradict \Eo. Note that a compiler that
translate all terms to unification variables would satisfy this
property, but wound certainly fail to verify \fullref{prop:fidelity}.

@@

We prove \fullref{prop:fidelity} in
\cref{sec:eta,sec:beta,sec:invariant1}
and \fullref{prop:simulation} only in \cref{sec:llam},
where links are not present yet.
In order to prove \cref{prop:simulation} in the presence of links
one needs to refine \cref{prop:fidelity} by relating the two
substitutions $\rho_p$ and $\sigma_p$: they are equivalent only by using decompilation to 
take into account terms in \linkStore.
Then, by induction
on the number of steps, \cref{prop:fidelity} implies \cref{prop:simulation}.
We discuss \fullref{prop:fidelity-recovery} in \cref{sec:beta}.
